Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-02 Manuscript/Manuscript.Rmd
Add some more refs to the intro

## Multilevel data

In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.^[Add IPDMA back in: patient level data from multiple studies.] Ignoring such structures may be harmful to the inferences and can yield biased results. On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account. Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa--for each of the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.

The variability due to clustering is often measured by means of the intra-class coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including separate intercepts for each cluster. Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters. Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

## Missing data

@@

The process of analyzing multilevel data is further complicated when not all observations are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness, see @resc13. We have visualized the difference between these types of missingness in Figure 1. The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n=5$ units and $p=3$ columns distributed over 3 clusters. Column `X1` is completely observed. Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why. GV: not necessary]

Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical. With the latter, the values would differ. The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern. ^[add missing data mechanisms here? GV: No, patterns are different from mechanisms]

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power. Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest. Especially the former is very generic and popular and is widely known as imputation. Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations^[add that JOMO is available in \pkg{mice} as well?]. Other useful packages for incomplete multilevel data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. So we won't go into detail for the different methods (and equations). Refer to @meng94, @audi18, and @grun18. This paper is just a software tutorial. We'll keep it practical.]

We structure this tutorial around three case studies:

- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MAR missingness in the outcome);

- `GJRM::hiv` (simulated patient data on HIV, with MNAR missingness in the outcome)

- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);^[add cross-level interaction?]

For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data. In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model. In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data. In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model. For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data:

A. Choose an analysis model (so the imputation model will be compatible with the analyses);
B. Evaluate the incomplete data;
C. Develop the imputation model(s);
D. Impute the missingness;
E. Evaluate the imputations.

@@

@@

>>>>

In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded. For example, students clustered within schools, or patients clustered within studies in individual patient data meta-analyses.  Analyzing such multilevel data requires specialized techniques that take the clustered structure into account, since ignoring it may be harmful to the statistical inferences and can yield biased results. Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.

The variability due to clustering is often measured by means of the intra-class coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including separate intercepts for each cluster. Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters. Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g. @hox17 and @jong21). There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

## Incomplete multilevel data

@@

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. We have visualized the difference between these types of missingness in Figure 1. The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns. Column `X1` is completely observed. Column `X2` is systematically missing and column `X3` is sporadically missing. Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical. With the latter, the values would differ. The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern.

Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness. Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following:

- Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries;
- Missing At Random (MAR), where the probability to be missing depends on observed information;
- Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94].
Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power. Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest. Especially the former is very generic and popular and is widely known as imputation. Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well). Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18. Note that this tutorial paper assumes a basic level of knowledge on multilevel models. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical.

To illustrate how to impute incomplete multilevel data, we structure this tutorial around three case studies:

- `mice::popmis` (simulated data on perceived popularity, $n = 199$ pupils across $N = 10$ classes, with univariate MAR missingness);
- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).

In each case study we highlight one aspect of the imputation workflow. With the `mice::popmis` data, we show different ways of designing an imputation model and what happens if you misspecify this model. With the `GJRM::hiv` data we extend the imputation model to include Heckman-type selection-inclusion methods. And with the `metamisc::impact` data we provide a example of multivariate missingness in real-world data. For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data: A. Choose an analysis model (so the imputation model will be compatible with the analyses); B. Evaluate the incomplete data; C. Develop the imputation model(s); D. Impute the missingness; E. Evaluate the imputations.

@@

Add: Assumes exchangability between units.

@@

Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.

The more the random effects are of interest, the more you need ml models.

### 2022-02-02 Manuscript/Manuscript.tex
Add some more refs to the intro

\hypertarget{multilevel-data}{%
\subsection{Multilevel data}\label{multilevel-data}}

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded.\footnote{Add IPDMA back in: patient
  level data from multiple studies.} Ignoring such structures may be
harmful to the inferences and can yield biased results. On the other
hand, analyzing such multilevel data requires specialized techniques
that take the clustered structure into account. Imagine a case where
cross-level interactions between unit-level variables and cluster-level
variables are present. The cluster to which a unit belongs may then
influence the unit-level observations--and vice versa--for each of the
units that make up the cluster \citep{hox17}. These relations can and
should be taken into account when developing analysis models for
multilevel data for the simple reason that groups of observations share
some common variance.

The variability due to clustering is often measured by means of the
intra-class coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure. Multilevel models typically accommodate for this variability
by including separate intercepts for each cluster. Such fixed effects
relieve the restriction imposed by single-level models: equal group
means across clusters. Additionally, there may be random predictor
effects and/or random error terms (residual error variances), see e.g.
\citet{hox17} and \citet{jong21}.\footnote{Add that heterogeneity refers
  to variability within clusters.} There are many names for models that
take clustering into account. Some popular examples are `multilevel
models', `hierarchical models', `mixed effect models' and `random effect
models'.

\hypertarget{missing-data}{%
\subsection{Missing data}\label{missing-data}}

@@

The process of analyzing multilevel data is further complicated when not
all observations are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Missingness in
multilevel data can therefore be categorized into two general patterns:
systematic missingness and sporadic missingness, see \citet{resc13}. We
have visualized the difference between these types of missingness in
Figure 1. The figure shows an \(n \times p\) set
\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
distributed over 3 clusters. Column \texttt{X1} is completely observed.
Column \texttt{X2} is systematically missing and column \texttt{X3} is
sporadically missing.\footnote{Explain why. GV: not necessary}

Systematic missingness can be further subdivided into unobserved
constants (i.e., the same value within clusters) and non-measured random
variables (which may differ per unit within clusters). In Figure 1, the
former would imply that the unobserved values for units 3 and 4 on
column \texttt{X2} are identical. With the latter, the values would
differ. The optimal strategy for dealing with the missingness may
therefore depend on the observed missing data pattern. \footnote{add
  missing data mechanisms here? GV: No, patterns are different from
  mechanisms}

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis (i.e., excluding all units with one
or more missing entries) can introduce bias in statistical inference and
lowers statistical power. Instead, the missingness should be
accommodated \emph{before} or \emph{within} the analysis of scientific
interest. Especially the former is very generic and popular and is
widely known as imputation. Imputing (i.e., filling in) the missing
values separates the missing data problem from the scientific problem:
missing data are replaced by plausible values whereafter the completed
data is analysed as if it were completely observed. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations\footnote{add that JOMO is available in
  \pkg{mice} as well?}. Other useful packages for incomplete multilevel
data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
  a software tutorial. We'll keep it practical.}

We structure this tutorial around three case studies:

\begin{itemize}
\item
  \texttt{mice::popmis} (simulated data on school kids, n = 199 units, N
  = 10 clusters, with MAR missingness in the outcome);
\item
  \texttt{GJRM::hiv} (simulated patient data on HIV, with MNAR
  missingness in the outcome)
\item
  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
  without \texttt{NA}s);\footnote{add cross-level interaction?}
\end{itemize}

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
show the advantages of including the multilevel structure of the data
into the imputation model. In the \texttt{metamisc::impact} data we'll
show how to induce missingness and solve it in real-world data. In the
\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
  methods exist already, but how to show that this is something new and
  exciting?} for imputing MNAR missingness according to the Heckman
model. For all case studies we discuss the nature of the incomplete
data, the imputation model(s), and evaluation of the imputed data:

A. Choose an analysis model (so the imputation model will be compatible
with the analyses); B. Evaluate the incomplete data; C. Develop the
imputation model(s); D. Impute the missingness; E. Evaluate the
imputations.

>>>>

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded. For example, students clustered within
schools, or patients clustered within studies in individual patient data
meta-analyses. Analyzing such multilevel data requires specialized
techniques that take the clustered structure into account, since
ignoring it may be harmful to the statistical inferences and can yield
biased results. Imagine a case where cross-level interactions between
unit-level variables and cluster-level variables are present. The
cluster to which a unit belongs may then influence the unit-level
observations--and vice versa for each of the units that make up the
cluster \citep{hox17}. These relations can and should be taken into
account when developing analysis models for multilevel data for the
simple reason that groups of observations share some common variance.

The variability due to clustering is often measured by means of the
intra-class coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure. Multilevel models typically accommodate for this variability
by including separate intercepts for each cluster. Such fixed effects
relieve the restriction imposed by single-level models: equal group
means across clusters. Additionally, models may include random effects
of predictors across clusters, and random error terms (heterogeneous
residual error variances; see e.g. \citet{hox17} and \citet{jong21}).
There are many names for models that take clustering into account. Some
popular examples are `multilevel models', `hierarchical models', `mixed
effect models' and `random effect models'.

\hypertarget{incomplete-multilevel-data}{%
\subsection{Incomplete multilevel
data}\label{incomplete-multilevel-data}}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Missingness in
multilevel data can therefore be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}. We have
visualized the difference between these types of missingness in Figure
1. The figure shows an \(n \times p\) set
\(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
\(N\) clusters, and \(p=3\) columns. Column \texttt{X1} is completely
observed. Column \texttt{X2} is systematically missing and column
\texttt{X3} is sporadically missing. Systematic missingness can be
further subdivided into unobserved constants (i.e., the same value
within clusters) and non-measured random variables (which may differ per
unit within clusters). In Figure 1, the former would imply that the
unobserved values for units 3 and 4 on column \texttt{X2} are identical.
With the latter, the values would differ. The optimal strategy for
dealing with the missingness may therefore depend on the observed
missing data pattern.

Another characteristic of the missing data to take into account in
analyses is the mechanism behind the missingness. Although the essence
of the true non-response mechanism may not be known, it can be inferred
or assumed to be one of the following:

\begin{itemize}
\tightlist
\item
  Missing Completely At Random (MCAR), where the probability to be
  missing is equal across all data entries;
\item
  Missing At Random (MAR), where the probability to be missing depends
  on observed information;
\item
  Missing Not At Random (MNAR), where the probability to be missing
  depends on unrecorded information, making the missingness
  non-ignorable \citep{rubi76, meng94}. Depending on the assumed
  missingness mechanism, missing data handling strategies may be more or
  less suitable, see e.g., \citet{yuce08} and \citet{hox15}.
\end{itemize}

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis (i.e., excluding all units with one
or more missing entries) can introduce bias in statistical inference and
lowers statistical power. Instead, the missingness should be
accommodated \emph{before} or \emph{within} the analysis of scientific
interest. Especially the former is very generic and popular and is
widely known as imputation. Imputing (i.e., filling in) the missing
values separates the missing data problem from the scientific problem:
missing data are replaced by plausible values whereafter the completed
data is analysed as if it were completely observed. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data mechanisms.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations exclusively (although the alternative,
joint modeling imputation for multilevel data or \pkg{jomo}
\citet{jomo}, has been implemented in \pkg{mice} as well). Other useful
resources for the analysis of incomplete multilevel data include the
\proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and
the empirical work by \citet{audi18} and \citet{grun18}. Note that this
tutorial paper assumes a basic level of knowledge on multilevel models.
We're providing an overview of implementations. It's up-to the reader to
decide which multilevel strategy suits their data. We won't go into
detail for the different methods (and equations). This paper is just a
software tutorial, so we'll keep it practical.

To illustrate how to impute incomplete multilevel data, we structure
this tutorial around three case studies:

\begin{itemize}
\tightlist
\item
  \texttt{mice::popmis} (simulated data on perceived popularity,
  \(n = 199\) pupils across \(N = 10\) classes, with univariate MAR
  missingness);
\item
  \texttt{GJRM::hiv} (simulated data on HIV diagnoses, \(n = 6,416\)
  patients across \(N = 9\) regions, with univariate MNAR missingness)
\item
  \texttt{metamisc::impact} (empirical data on traumatic brain injuries,
  \(n = 11,022\) patients across \(N = 15\) studies, without
  \texttt{NA}s).
\end{itemize}

In each case study we highlight one aspect of the imputation workflow.
With the \texttt{mice::popmis} data, we show different ways of designing
an imputation model and what happens if you misspecify this model. With
the \texttt{GJRM::hiv} data we extend the imputation model to include
Heckman-type selection-inclusion methods. And with the
\texttt{metamisc::impact} data we provide a example of multivariate
missingness in real-world data. For all case studies we discuss the
nature of the incomplete data, the imputation model(s), and evaluation
of the imputed data: A. Choose an analysis model (so the imputation
model will be compatible with the analyses); B. Evaluate the incomplete
data; C. Develop the imputation model(s); D. Impute the missingness; E.
Evaluate the imputations.
